The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-14 script/de/chapters/numbertheory.tex
CT script: Sage parts finished in (en); translation to German almost finished

The Multiplikationstabelle~\ref{mulmod17} (von Seite \pageref{SrcArith1a})
for $a \times i \pmod{m}$, mit $m = 17$, $a=5$ and $a=6$, und $i$ von $0$ bis $16$
kann mit folgenden Sage-Befehlen berechnet werden:

@@

The other multiplication table examples modulo $13$ (table~\ref{mulmod13})
and modulo $12$ (table~\ref{mulmod12}) on page \pageref{SrcArith1b}
can similarly be computed by replacing {\tt m = 17}
with {\tt m = 13} and {\tt m = 12} respectively.

@@

\newpage
\newpage
\noindent Mit unserer selbst erstellten Funktion \verb!enum_PrimitiveRoots_of_an_Integer!
kann man alle Primitivwurzeln einer Primzahl $p$ finden.

@@

\newpage
\newpage
\noindent Eine leicht geänderte Fassung unserer Funktion
\verb!count_PrimitiveRoots_of_a_PrimesRange! wird nun benutzt, um eine
Liste (Datenbank) aller Primitivwurzeln für alle Primzahlen zwischen 1 und 100.000
zu erstellen.

@@

\newpage
\newpage
\noindent Diese Datei ``primroots.dat'' wurde dann benutzt, um mit dem folgenden Code
drei Grafiken zu erstellen.

@@

% To enforce that the next chapter starts at the beginning of a new page after
% the freely placed graphics, it is necessary to have TWO newpages and in between
% something to be printed (here invisible blanks).
% \newpage
% $ $

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% \vskip +12pt
\newpage
\hypertarget{NumberTheory_Appendix_F}{}  %\hypertarget{AppendixListAndDef}{}
\section{Anhang: Liste der in diesem Kapitel formulierten Definitionen und Sätze}
\label{l:NumberTheory_Appendix_F}{}  %\label{l:AppendixListAndDef}{}

>>>>

The Multiplikationstabelle~\ref{mulmod17} (von Seite \pageref{SrcArith1a})
für $a \times i \pmod{m}$, mit $m = 17$, $a=5$ and $a=6$, und $i$ von $0$ bis $16$
kann mit folgenden Sage-Befehlen berechnet werden:

@@

@@

\clearpage
\noindent Mit unserer selbst erstellten Funktion \verb!enum_PrimitiveRoots_of_an_Integer!
kann man alle Primitivwurzeln einer Primzahl $p$ finden.

@@

\newpage
\noindent Eine leicht geänderte Fassung unserer Funktion
\verb!count_PrimitiveRoots_of_a_PrimesRange! wird nun benutzt, um eine
Liste (Datenbank) aller Primitivwurzeln für alle Primzahlen zwischen 1 und 100.000
zu erstellen.

@@

\newpage
\noindent Diese Datei ``primroots.dat'' wurde dann benutzt, um mit dem folgenden Code
drei Grafiken zu erstellen.

@@

% To enforce that the next chapter starts at the beginning of a new page after
% the freely placed graphics, it was necessary to have TWO newpages and in between
% something to be printed (here invisible blanks).
% \newpage
% $ $
% I guess \clearpage is preferable.

@@

% ---------------------------------------------------------------------------
\newpage
\hypertarget{NumberTheory_Sage_Number-of-RSA-keys}{}
\subsection{Wie viele RSA-Schlüssel gibt es innerhalb eines Modulo-Bereiches?}
\label{l:NumberTheory_Sage_Number-of-RSA-keys}{}

Die RSA-Verschlüsselung wurde beschrieben in Abschnitt \ref{RSA} (``\nameref{RSA}'').
Schritt 1 bis 3 definieren die Schlüsselerzeugung, Schritt 4 und 5 stellen die
eigentliche Verschlüsselung dar:
\begin{itemize}
  \item[{\bf 1.}] Wähle zwei unterschiedliche Primzahlen $p$ and $q$
                  und berechne $n = p*q$.\\
                  Der Wert $n$ wird RSA-Modul genannt.

  \item[{\bf 2.}] Wähle eine zufälliges $e \in \{2, \cdots, n-1\}$ so dass gilt: \\
                  $e$ ist relativ prim zu $J(n) = (p-1)*(q-1)$. \\
                  Danach kann man $p$ und $q$ \glqq wegwerfen\grqq.

  \item[{\bf 3.}] Wähle $d \in \{1, \cdots, n-1\}$ mit $e*d \equiv 1  
                  {\rm ~(mod~} J(n))$, d.h. $d$ ist die multiplikative Inverse
                  von $e$ modulo $J(n)$. Danach kann man $J(n)$ \glqq wegwerfen\grqq.
    \begin{itemize}
      \item[] $\rightarrow (n, e)$ ist der öffentliche Schlüssel $P$.
      \item[] $\rightarrow (n, d)$ ist der private Schlüssel $S$ (nur $d$ muss man geheim halten).
    \end{itemize}

  \item[{\bf 4.}] Zur Verschlüsselung wird die Nachricht als (binäre) Ziffernfolge
                  geschrieben. Diese Nachricht wird dann so in gleich lange Teilfolgen
		  aufgeteilt, dass jeder Teil der Zahl kleiner als $n$ ist.

  \item[{\bf 5.}] Vorgang der Verschlüsselung auf dem Klartext (bzw. auf Teilen davon)
                  $M \in \{1, \cdots, n-1\}$:
                  $$C = E ( (n, e); M ) := M^e {\rm ~(mod~} n).$$
\end{itemize}

Standardmäßig versucht man, einen mit RSA verschlüsselten Geheimtext $C$ dadurch zu knacken,
dass man den öffentlichen Schlüssel des Empfängers betrachtet und versucht, $n$ zu
faktorisieren. Hat man das erreicht, geht man wieder durch die Schritte 2 und 3 und
berechnet deb privaten Schlüssel $e$, den man entschlüsseln des Geheimtextes braucht.

Gemäß dem \glqq Primzahlsatz\grqq (beschrieben in Abschnitt \ref{thm-pz-pi-x}
\glqq \nameref{thm-pz-pi-x}\grqq) bewegt sich die Anzahl der Primzahlen
$PI(x)$ asymptotisch gegen  $x / ln(x)$.
Für ein gegebenes $n$ gibt es also ca. $n / ln(n)$ mögliche Werte für die
Primzahl $p$.

\noindent Will man nicht faktorisieren, sondern stellt sich eine Frage ähnlich
wie bei den klassischen Verschlüsselungsverfahren, sondern kann man herausfinden wollen:
Wie viele verschiedene private Schlüssel $(n, d)$ gibt es für einen bestimmten
Bereich der Schlüsselgröße $n \in \{a, b\}$?

\noindent Das Sage-Besipielprogramm~\ref{nt_sagesample_Count_RSA_Keys} unten definiert
die Funktion \verb#count_Number_of_RSA_Keys#, die diese Frage konkret beantworten kann
(wenn der Modulus nicht zu groß ist).\footnote{%
a) Der Aufruf \verb#sage: count_Number_of_RSA_Keys(100, 1000)# bedeutet, dass man
das Intervall $[100, 1000]$ für $n$ betrachtet.
$n$ war definiert durch die beiden Primzahlen $p, q: n = p*q$.
Daher kann eine Primzahl höchstens dern Wert $500$ annehmen, weil $2 * 500 =1000$.\\
Die Anzahl möglicher Primzahl-Kombinationen beträgt: $comb = 258$.\\
Die Anzahl der Primzahlen im gegebenen Bereich beträgt: $143$.\\
Die Anzahl der privaten Schlüssel beträgt: $34.816$.

\noindent b) Der Aufruf \verb#sage: count_Number_of_RSA_Keys(100, 100, True)#
hat die folgende Ausgabe:\\
   - Number of private keys for modulus in a given range: 0\\
   - Number of primes in a given range: 0\\
   Der Grund dafür ist, dass wir damit nur $n=100$ betrachten, und $100$ ist nicht
   semi-prim\index{Primzahl!semi-prim}\index{Primzahl!Halbprimzahl},
   d.h. $100$ ist nicht das Produkt von genau zwei Primzahlen.
}

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
def count_Number_of_RSA_Keys(start, end, Verbose=False):
      r"""
      How many private RSA keys (n,d) exist, if only modulus N is given, and start <= N <= end?
        (prime_range(u,o) delivers all primes >=u und < o).
      """
      a = start
      b = end
      s = 0
      comb = 0
      for p in prime_range(1, b/2+1):
          for q in prime_range(p + 1, b/2+1):
              if a <= p * q and p * q <= b:
                  comb = comb +1
                  s = s + (euler_phi(euler_phi(p * q))-1)
                  if Verbose:
                      print "p=%s, " % p + "q=%s, " % q + "s=%s" % s
      print "Number of private keys for modulus in a given range: %s" % s + " (comb=%s), " % comb

      # Just for comparison: How many primes are in this range?
      s = 0
      for p in prime_range(a, b+1):
          if Verbose:
              print "a=%s, " % a + "b=%s, " % b + "p=%s" % p
          s = s + 1
      print "Number of primes in a given range: %s" % s

print "\n\nDD_Start -- Testcases for count_Number_of_RSA_Keys(start, end)"
print "\n-----------Testcase: (100, 1000) [Should deliver 34.816]"
time count_Number_of_RSA_Keys(100, 1000)
print "\n-----------Testcase: (100, 107, True) [Should deliver 23]"
time count_Number_of_RSA_Keys(100, 107, True)
u = 10^3; o = 10^4;
print "\n-----------Testcase: (%s, " % u + "%s) [Should deliver 3.260.044]" % o
time count_Number_of_RSA_Keys(u, o)

OUTPUT:
DD_Start -- Testcases for count_Number_of_RSA_Keys(start, end)

-----------Testcase: (100, 1000) [Should deliver 34.816]
Number of private keys for modulus in a given range: 34816 (comb=258),
Number of primes in a given range: 143
Time: CPU 0.03 s, Wall: 0.04 s

-----------Testcase: (100, 107, True) [Should deliver 23]
p=2, q=53, s=23
Number of private keys for modulus in a given range: 23 (comb=1),
a=100, b=107, p=101
a=100, b=107, p=103
a=100, b=107, p=107
Number of primes in a given range: 3
Time: CPU 0.00 s, Wall: 0.00 s

-----------Testcase: (1000, 10000) [Should deliver 3.260.044]
Number of private keys for modulus in a given range: 3260044 (comb=2312),
Number of primes in a given range: 1061
Time: CPU 0.63 s, Wall: 0.66 s
\end{Verbatim}
\caption{Wie viele RSA-Schlüssel gibt es, wenn man den Bereich für die Schlüsselgröße n kennt?}
\label{nt_sagesample_Count_RSA_Keys}
\end{sagecode}

\noindent Weil es mehr private Schlüssel $(n, d)$ innerhalb eines größeren
Bereiches von Werten für $n$ gibt, ist sogar das Brute-force-Faktorisieren
viel effizienter als alle möglichen Schlüssel durchzuprobieren.

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\hypertarget{NumberTheory_Appendix_F}{}  %\hypertarget{AppendixListAndDef}{}
\section{Anhang: Liste der in diesem Kapitel formulierten Definitionen und Sätze}
\label{l:NumberTheory_Appendix_F}{}  %\label{l:AppendixListAndDef}{}

### 2009-08-14 script/de/chapters/what-is-sage.tex
Language de: Add an appendix on using Sage with this script; TODO: translate
incorporated user feedback, removed some \newpage
CT script: Sage parts finished in (en); translation to German almost finished

% ---------------------------------------------------------------------------
\chapter*{Using Sage with this Script}
\addcontentsline{toc}{chapter}{Using Sage with this Script}

This script includes numerous code samples using Sage. Sage is an open
source computer algebra system that supports teaching, study and
research in mathematics.  It combines numerous high-quality open
source packages and provides access to their functionalities via a
common interface, namely, a Python based programming language.  Sage
can be used as a powerful desktop calculator, as a tool to help
undergraduate students study mathematics, or as a programming
environment for prototyping algorithms and research in algorithmic
aspects of mathematics.  With respect to studying cryptography, Sage
modules in the directory  \url{SAGE_ROOT/devel/sage-main/sage/crypto}
can be used to complement a first course in cryptography.  David
Kohel's notes at
\begin{center}
  \url{http://www.sagemath.org/library/crypto.pdf}
\end{center}
can be used to teach such a course that incorporates Sage.

\section*{Sage user interfaces}

Sage is available free of charge and can be downloaded from the
following website:
\begin{center}
  \url{http://www.sagemath.org} \\
\end{center}
The default interface to Sage is command line based, as shown in
Figure~\ref{fig:sage_cmd_interfaces}. However, there is a
graphical user interface to the software as well in the form of the
Sage notebook (see Figure~\ref{fig:sage_gui_interfaces}). We can even
use Sage online at
\begin{center}
\url{http://www.sagenb.org}
\end{center}
without having to install Sage locally. Sage runs under many popular
Linux distributions, Mac OS X, and Windows. For the Windows platform,
a complete distribution of Sage currently only runs as a VMware
image. However, a full native port of Sage to Windows is currently in
progress.

\begin{figure}[!htpb]
\centering
\includegraphics[scale=0.5]{figures/sage-cmd}
\caption{Sage command line interface.}
\label{fig:sage_cmd_interfaces}
\end{figure}

\begin{figure}[!htpb]
\centering
\includegraphics[scale=0.4]{figures/sage-gui}
\caption{Sage notebook interface.}
\label{fig:sage_gui_interfaces}
\end{figure}

\newpage

\section*{Getting help with using Sage}

Upon loading Sage from the command line, we are presented with
something similar to the following:
%
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
----------------------------------------------------------------------
| Sage Version 4.0.2, Release Date: 2009-06-18                       |
| Type notebook() for the GUI, and license() for information.        |
----------------------------------------------------------------------
sage:
\end{Verbatim}
%
Plenty of help is provided in the form of the official Sage
documentation that is distributed with every release of Sage~(see
Figure~\ref{fig:sage_standard_doc}).
%
\begin{figure}[!htpb]
\centering
\includegraphics[scale=0.4]{figures/sage-online-doc}
\caption{The Sage standard documentation.}
\label{fig:sage_standard_doc}
\end{figure}
%
The official Sage standard documentation includes the following documents:

@@

>>>>

% ---------------------------------------------------------------------------
\newpage
\hypertarget{appendix-using-sage}{}
\section{Kurzeinführung in das Computeralgebrasystem Sage}
\label{s:appendix-using-sage}
\index{Sage}
\index{Sage!Programmbeispiele}

Dieses Skript enthält zahlreiche Programmbeispiele, die Sage nutzen. Sage ist
ein Open-Source Computer-Algebra-System (CAS), das für Lehre, Studium und Forschung
im Mathematik eingesetzt wird.
Sage kombiniert viele hochwertige Open-Source-Packete\footnote{%
Einen Eindruck von der Größe von Sage erhält man, wenn man es selbst compiliert:
Die heruntergeladenen Sourcen von Sage 4.1 brauchten zur Compilierung auf einem
durchschnittlichen Linux-PC rund 5 h (inklusive aller Bibliotheken).
Danach nahm es 1,8 GB Plattenplatz ein.
% TODO JOERG
% Ich habe noch kash3-2008-07-31.spkg nachinstalliert, dieses Paket wird für Frau Massierers ECC Notebooks benötigt.
}
und liefert den Zugang zu deren Funktionalität über ein gemeinsames, auf der
Programmiersprache Python basierendes Interface\footnote{%
Es gibt auch ein relativ einfaches Interface für die Sprache C, genannt Cython,
mit der man eigene Funktionen in Sage stark beschleunigen kann.\\
Siehe \url{http://openwetware.org/wiki/Open_writing_projects/Sage_and_cython_a_brief_introduction}.
}.

\noindent Sage kann man wie andere CAS auch uf vielfältige Weise nutzen:
als mächtigen Taschenrechner; als Tool, das Studenten beim Mathematikstudium hilft;
oder als Programmier-Umgebung, um Algorithmen zu prototypen oder um Forschung im
Bereich der algorithmischen Aspekte der Mathematik zu betreiben.

\noindent Einen schnellen Eindruck von Sage kann man gewinnen mit der
\glqq Einladung zu Sage\grqq von David Joyner\footnote{%
\url{http://sage.math.washington.edu/home/wdj/teaching/calc1-sage/an-invitation-to-sage.pdf}\\
(Letztes Update 2009).}.

\noindent Die offizielle Sage Online-Dokumentation\footnote{%
  Die entsprechenden offiziellen PDF-Dokuments können herunter geladen werden von\\
  \url{http://www.sagemath.org/help.html}
} ist verfügbar unter: \url{http://www.sagemath.org/doc}.

\noindent Es gibt inzwischen viele PDF- oder HTML-Dokumente über Sage:
Ein guter Startpunkt ist die \glqq Bibliothek\grqq von der Sage-Hauptseite,
das \glqq Dokumentationsprojekt\grqq und das \glqq Lehrmaterial\grqq\footnote{%
- \glqq Bibliothek\grqq: {\centering \url{http://www.sagemath.org/library/index.html}},\\
- \glqq Dokumentationsprojekt\grqq:
  {\centering \url{http://wiki.sagemath.org/DocumentationProject}},\\
- \glqq Lehrmaterial\grqq: {\centering \url{http://wiki.sagemath.org/Teaching_with_SAGE}}.
}.
xxxxxxxxxxxxxxxxx
\noindent With respect to studying cryptography, Sage modules can be used
to complement a first course in cryptography\footnote{%
- See the module sources in the directory \url{SAGE_ROOT/devel/sage-main/sage/crypto}.\\
- Overview, what crypto currently is in Sage:\\
~ ~ \url{http://www.sagemath.org/doc/reference/sage/crypto/}\\
- Discussions about the teaching related aspect of development crypto in Sage:\\
~ ~ \url{http://groups.google.com/group/sage-devel/browse_thread/thread/c5572c4d8d42d081}
}.

\noindent Especially David Kohel's notes at\\
{\centering \url{http://www.sagemath.org/library/crypto.pdf} }
or the same eventually newer at\\
{\centering \url{http://sage.math.washington.edu/home/wdj/teaching/kohel-crypto.pdf} }\\
can be used to teach such a course that incorporates Sage.

% ---------------------------------------------------------------------------
\section*{Sage user interfaces}
Sage is available free of charge and can be downloaded from the
following website:
\begin{center}
  \url{http://www.sagemath.org} \\
\end{center}
The default interface to Sage is command line based, as shown in
Figure~\ref{fig:sage_cmd_interfaces}. However, there is a
graphical user interface to the software as well in the form of the
Sage notebook (see Figure~\ref{fig:sage_gui_interfaces}). We can even
use Sage online at
\begin{center}
\url{http://www.sagenb.org}
\end{center}
without having to install Sage locally. Sage runs under many
Linux distributions, Mac OS X, and Windows. For the Windows platform,
a complete distribution of Sage currently only runs as a VMware
image. However, a full native port of Sage to Windows is currently in
progress.

\begin{figure}[!htpb]
\centering
\includegraphics[scale=0.6]{figures/sage-cmd}
\caption{Sage command line interface}
\label{fig:sage_cmd_interfaces}
\end{figure}

\begin{figure}[!htpb]
\centering
\includegraphics[scale=0.4]{figures/sage-gui}
\caption[Sage notebook interface]{Sage notebook interface\footnotemark}
\label{fig:sage_gui_interfaces}
\end{figure}

\footnotetext{%
Enter notebook() at the Sage prompt, and then start your favorite browser
(Iceweasel, Firefox, IE, ...) e.g. with the URL \url{http://localhost:8000}.
}

% ---------------------------------------------------------------------------
\newpage
\section*{Getting help with using Sage}

Upon loading Sage from the command line, we are presented with
something similar to the following:
%
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
mnemonic:~$ sage
----------------------------------------------------------------------
| Sage Version 4.1, Release Date: 2009-07-09                         |
| Type notebook() for the GUI, and license() for information.        |
----------------------------------------------------------------------

sage: help
Type help() for interactive help, or help(object) for help about object.
sage:
sage:
sage: help()

Welcome to Python 2.6!  This is the online help utility.

If this is your first time using Python, you should definitely check out
the tutorial on the Internet at http://docs.python.org/tutorial/.

Enter the name of any module, keyword, or topic to get help on writing
Python programs and using Python modules.  To quit this help utility and
return to the interpreter, just type "quit".

To get a list of available modules, keywords, or topics, type "modules",
"keywords", or "topics".  Each module also comes with a one-line summary
of what it does; to list the modules whose summaries contain a given word
such as "spam", type "modules spam".
\end{Verbatim}
%
Plenty of help is provided in the form of the official Sage
documentation that is distributed with every release of Sage~(see
Figure~\ref{fig:sage_standard_doc}).
%
\begin{figure}[!htpb]
\centering
\includegraphics[scale=0.4]{figures/sage-online-doc}
\caption{The Sage standard documentation}
\label{fig:sage_standard_doc}
\end{figure}
%
The official Sage standard documentation includes the following documents:

@@

% ---------------------------------------------------------------------------
\newpage
\section*{Some examples using the built-in mathematical functions in Sage}

Here are a few little examples\footnote{%
The examples are from the blog of Prof. Alasdair McAndrew, Victoria University,\\
\url{http://amca01.wordpress.com/2008/12/19/sage-an-open-source-mathematics-software-system}}
(all in console mode, for ease) to see what you can do with Sage:

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
# * Calculus:
    sage: x=var('x')
    sage: p=diff(exp(x^2),x,10)*exp(-x^2)
    sage: p.simplify_exp()
     1024 x^10 + 23040 x^8 + 161280 x^6 + 403200 x^4 + 302400 x^2 + 30240

# * Linear Algebra:
    sage: M=matrix([[1,2,3],[4,5,6],[7,8,10]])
    sage: c=random_matrix(ZZ,3,1);c
     [ 7 ]
     [-2 ]
     [-2 ]
    sage: b=M*c
    sage: M^-1*b
     [ 7 ]
     [-2 ]
     [-2 ]

# * Number theory:
    sage: p=next_prime(randint(2^49,2^50));p
      1022095718672689
    sage: r=primitive_root(p);r
      7
    sage: pl=log(mod(10^15,p),r);pl
      1004868498084144
    sage: mod(r,p)^pl
      1000000000000000

# * Finite Fields:
    sage: F.<x>=GF(2)[]
    sage: G.<a>=GF(2^4,name='a',modulus=x^4+x+1)
    sage: a^2/(a^2+1)
      a^3 + a
    sage: a^100
      a^2 + a + 1
    sage: log(a^2,a^3+1)
      13
    sage: (a^3+1)^13
      a^2
\end{Verbatim}
\caption{Some small general samples from different areas in mathematics}
\end{sagecode}

% ---------------------------------------------------------------------------
\newpage
\section*{Writing code samples with Sage}

At the beginning using a CAS (computer algebra system) you type in the single commands
on the command line as in the above example\footnote{%
  The standard way for presenting Sage code starts the lines with ``sage:'' and ``...''.
  \begin{Verbatim}%
  [fontsize=\footnotesize,fontshape=tt]
  sage: m = 11
  sage: for a in xrange(1, m):
  ....:     print [power_mod(a, i, m) for i in xrange(1, m)]
  ....:
  \end{Verbatim}

  \noindent This script usually uses the above convention for presenting Sage code,
  if the code doesn't come from a Sage script. When people copy and paste the Sage code
  from the script, they should leave out "sage:" and "..."from the script (nevertheless
  in most cases the command prompt can deal with this prefixes correctly).}.

But if you develop your own functions, modify them and call them, then it is much easier
to do the development in your own editor, save it to a script file and execute the functions
non-iteractively on the commandline manually or even from within the script file.
Both is done in the examples in
chapter \ref{PaP_Sage_samples} (``\nameref{PaP_Sage_samples}'') and in
chapter \ref{NumberTheory_Appendix_E} (``\nameref{NumberTheory_Appendix_E}'').

\noindent To program and test Sage code there are two useful commands: load() and attach()\footnote{%
See Sage tutorial about Programming, chapter ``Loading and Attaching SAGE files'',\\
\url{http://www.sagemath.org/doc/tutorial/programming.html\#loading-and-attaching-sage-files}.}.\\
Suppose you have a function definition like this:
\begin{Verbatim}%
   [fontsize=\footnotesize,fontshape=tt]
   def function(var1):
       r"""
       DocText.
       """
       ...
       return (L)
\end{Verbatim}
\noindent which has been saved to the file \texttt{primroots.sage}.\footnote{%
Note: don't use white spaces in your file name.}

\noindent To load this function into Sage, use load() as follows:

\texttt{sage: load primroots.sage}

\noindent and you can then proceed to use on the command line anything that is
defined in that Sage script\footnote{%
Its recommended that your Sage script has the file extension
``.sage'', instead of ``.py''. With a Sage script whose file name ends in
``.sage'', when you load it into Sage then the default Sage environment
is also loaded to make sure that it works as if you have defined your
function from the Sage command line. This also applies if you run the
script from a bash shell using ~~ \texttt{\$ sage primroots.sage}.

If you run your script as above, then Sage first parses your script,
writes it to another file called ``primroots.py'' (note the ``.py''
extension), adds all necessary variables to "primroots.py" as well as
writing any import statements to that file. That way, your Sage script
is executed as if you had typed the definitions in your script to the
Sage command line.}.

Sometimes we also want to edit a Sage script and reload the
content of the changed script into Sage again. In that case, you can
use the command attach(). Going back to our example of the above
function definition, say now you attach it to a Sage session as
follows:

\texttt{sage: attach primroots.sage}

Now edit the Sage script in a text editor, but don't exit Sage.
After you saved it within your text editor, the changed function
definition is reloaded into the running Sage session. This reloading is
done automatically for you, provided that all changes to your script
have been saved. You can think of the command attach() as a way of
telling Sage to watch for all changes to a file, and reloading the
file again once Sage notices that there have been changes. With this
command, you don't have to copy and paste between your text editor and
the Sage command line interface.

Here is a picture of Sage code in the editor GVIM with activated code highlighting
~(see Figure~\ref{fig:sage-highlighted-code-in-editor}).
%
\begin{figure}[!htpb]
\centering
\includegraphics[scale=0.7]{figures/sage-highlighted-code-in-editor}
\caption{Sage sample shown in an editor with code highlighting}
\label{fig:sage-highlighted-code-in-editor}
\end{figure}

\vspace{30pt}
To move quickly to the Sage code examples in this script, either look in the
index at \verb#Sage -> Code examples# or have a look at the
``\nameref{sc:List-of-Sage-Code-Examples}''.

### 2009-08-14 script/en/chapters/numbertheory.tex
CT script: Sage parts finished in (en); translation to German almost finished

\newpage
\newpage
\noindent Using our custom-defined function \verb!enum_PrimitiveRoots_of_an_Integer!,
we can find all primitive roots of one prime integer $p$.

@@

\newpage
\newpage
\noindent A slightly modified version of our function
\verb!count_PrimitiveRoots_of_a_PrimesRange!, is used to generate
a database of all primitive roots of all primes between 1 and 100,000.

@@

\newpage
\newpage
\noindent This database file ``primroots.dat'' was used then to create three graphics using the following code.

@@

% To enforce that the next chapter starts at the beginning of a new page after
% the freely placed graphics, it is necessary to have TWO newpages and in between
% something to be printed (here invisible blanks).
% \newpage
% $ $

@@

\vskip +10 pt 
\hypertarget{AppArith4b}{%
\noindent {\bf Example on page~\pageref{SrcArith4b}:}} \\
The factorization of $J(256027) = 255016 = 2^3 * 127 * 251$ can be
calculated using Sage as follows:

@@

The RSA encryption procedure was described in chapter \ref{RSA} (``\nameref{RSA}'').
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption:
\begin{itemize}
  \item[{\bf 1.}] Select two distinct random prime numbers $p$ and $q$
                  and calculate $n = p*q$.\\
                  The value $n$ is called the RSA modulus.

@@

The default way to crack a given RSA ciphertext $C$ would be to use the public key
and to try to factorize $n$. Then you can go through the steps 2 and 3 and calculate
the private key $e$, which is normally used to decrypt a ciphertext.

According to the ``prime number theorem''
(described in chapter \ref{thm-pz-pi-x} ``\nameref{thm-pz-pi-x}'')
the number of prime numbers $PI(x)$ is asymptotic to  $x / ln(x)$.
So in a given range $[x,y]$ there are about $ (y / ln(y)) - (x / ln(x))$
primes. If you have a given $n$ then there are about $n / ln(n)$ different possible
values for the prime $p$ (not considering the more specific requirements for selecting
appropriate values for $p$ as described above).

\noindent If you would not use factorization but ask the question like in classic encryption:
How many possible private keys $(n, d)$ are there for a given $n$ (this means only a part
of the public key $(n, e)$ is known).

\noindent Below is Sage source code~\ref{nt_sagesample_Count_RSA_Keys} which can answer
this question concretely (if the modulus is not too big).

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
def count_Number_of_RSA_Keys(start, end, Verbose=False):
      r"""
      How many private RSA keys (n,d) exist, if only modulus N is given, and start <= N <= end?
        (prime_range(u,o) delivers all primes >=u und < o).
      """
      a = start
      b = end
      s = 0
      comb = 0
      for p in prime_range(1, b/2+1):  # b or b/2+1
          for q in prime_range(p + 1, b/2+1):  # b or b/2+1
              if a <= p * q and p * q <= b:
                  comb = comb +1
                  s = s + (euler_phi(euler_phi(p * q))-1)
                  if Verbose:
                      print "p=%s, " % p + "q=%s, " % q + "s=%s" % s
      print "Number of private keys for modulus in a given range: %s" % s + " (comb=%s), " % comb

@@

-----------Testcase: (1000, 10000) [Should deliver 3.260.044]
Number of private keys for modulus in a given range: 3260044 (comb=2312),
Number of primes in a given range: 1061
Time: CPU 0.63 s, Wall: 0.66 s
\end{Verbatim}
\caption{How many RSA keys are there knowing the public key?}
\label{nt_sagesample_Count_RSA_Keys}
\end{sagecode}

>>>>

\newpage
\noindent Using our custom-defined function \verb!enum_PrimitiveRoots_of_an_Integer!,
we can find all primitive roots of one prime integer $p$.

@@

\newpage
\noindent A slightly modified version of our function
\verb!count_PrimitiveRoots_of_a_PrimesRange!, is used to generate
a database of all primitive roots of all primes between 1 and 100,000.

@@

\newpage
\noindent This database file ``primroots.dat'' was used then to create three graphics using the following code.

@@

@@

\vskip +10 pt 
\hypertarget{AppArith4b}{%
\noindent {\bf Example on page~\pageref{SrcArith4b}:}} \\

The factorization of $J(256027) = 255016 = 2^3 * 127 * 251$ can be
calculated using Sage as follows:

@@

The RSA encryption procedure was described in section \ref{RSA} (``\nameref{RSA}'').
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption:
\begin{itemize}
  \item[{\bf 1.}] Select two distinct random prime numbers $p$ and $q$
                  and calculate $n = p*q$.\\
                  The value $n$ is called the RSA modulus.

@@

The default way to crack a given RSA ciphertext $C$ would be to use the public key
of the recipient and to try to factorize $n$. Then you can go through the steps 2 and 3
and calculate the private key $e$, which is normally used to decrypt a ciphertext.

According to the ``prime number theorem''
(described in section \ref{thm-pz-pi-x} ``\nameref{thm-pz-pi-x}'')
the number of prime numbers $PI(x)$ is asymptotic to  $x / ln(x)$.
%% So in a given range $[x,y]$ there are about $ (y / ln(y)) - (x / ln(x))$ primes.
If you have a given $n$ then there are about $n / ln(n)$ different possible
values for the prime $p$. %% (not considering the more specific requirements for selecting
%% appropriate values for $p$ as described above).

\noindent If you don't want to use factorization but ask the question like in classic
encryption, you may want to find out:
How many possible private keys $(n, d)$ are there for a given key size range
$n \in \{a, b\}$? %% (this means, only a part of each public key $(n, e)$ is known).

\noindent Sage source code~\ref{nt_sagesample_Count_RSA_Keys} below defining the
function \verb#count_Number_of_RSA_Keys# can answer this question concretely
(if the modulus is not too big).\footnote{%
a) Calling \verb#sage: count_Number_of_RSA_Keys(100, 1000)# means to consider the interval
$[100, 1000]$ for $n$. $n$ is defined by the two primes $p, q: n = p*q$.
So the primes can have the maximal value $500$ because $2 * 500 =1000$.\\
The number of possible combinations of primes is $comb = 258$.\\
The number of primes in the given range is $143$.\\
The number of private keys is $34,816$.

\noindent b) Calling \verb#sage: count_Number_of_RSA_Keys(100, 100, True)# has the following output:\\
   - Number of private keys for modulus in a given range: 0\\
   - Number of primes in a given range: 0\\
   The reason for that is, that now we only consider $n=100$, and $100$ is not
   semi-prime\index{Prime number!semi-prime}\index{Prime number!half prime},
   this means $100$ is not the product of only two primes.
}

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
def count_Number_of_RSA_Keys(start, end, Verbose=False):
      r"""
      How many private RSA keys (n,d) exist, if only modulus N is given, and start <= N <= end?
        (prime_range(u,o) delivers all primes >=u und < o).
      """
      a = start
      b = end
      s = 0
      comb = 0
      for p in prime_range(1, b/2+1):
          for q in prime_range(p + 1, b/2+1):
              if a <= p * q and p * q <= b:
                  comb = comb +1
                  s = s + (euler_phi(euler_phi(p * q))-1)
                  if Verbose:
                      print "p=%s, " % p + "q=%s, " % q + "s=%s" % s
      print "Number of private keys for modulus in a given range: %s" % s + " (comb=%s), " % comb

@@

-----------Testcase: (1000, 10000) [Should deliver 3.260.044]
Number of private keys for modulus in a given range: 3260044 (comb=2312),
Number of primes in a given range: 1061
Time: CPU 0.63 s, Wall: 0.66 s
\end{Verbatim}
\caption{How many RSA keys are there knowing the range of the public key n?}
\label{nt_sagesample_Count_RSA_Keys}
\end{sagecode}
